Architectures for workflow integration in science gateways

### 2016-04-19 paper.tex
Updates in Section 3

The architectures are diagramed in Figure~\ref{fig:architectures}
using the graphical notations shown in
Figure~\ref{fig:notations}. Architectures are described from their
main software \emph{components} and \emph{interactions}. Software
components include science gateway and infrastructure in all
architectures, and workflow service, workflow pool and agent are
involved in some architectures. The workflow engine itself is
represented by a specific symbol. Software interactions among these
components are described in the next Section. \emph{Abstract}
interactions are specific types of interactions that may be
implemented by various different software interactions. An
architecture that has an abstract interaction is an abstract
architecture. Red color is used to represent the components and
interactions that are specific to workflow execution, i.e. that are
not involved in any other process. Tasks refer to computing tasks that
are created by the workflow engine and executed on the
infrastructure. Data represents any type of file, or database that is
involved in the workflow execution and stored on the infrastructure
while the workflow executes. It does not cover workflow parameters
such as strings, numbers, etc. The remainder of this section describes
the different types of interactions involved in the architectures, and
details each architecture.

@@

\item Workflow integration: consists in adding a new workflow to the
  system so that users can execute it. It is triggered by an
  administrator of the science gateway and it results in an interface,
  for instance a web form, where users can enter the parameters of the
  workflow to be executed. The interaction has two aspects:
  (\texttt{a$_1$}) the programs used in the workflow are installed on
  the infrastructure, which may or may not require administrative
  privileges on the infrastructure; (\texttt{a$_2$}) the workflow has
  to be configured in the science gateway so that it becomes available
  to users. Note that integrating a workflow is not the same process
  as integrating a workflow \emph{engine}.
\item Task control: operations to manage tasks on the infrastructure,
  including: authentication, submission, monitoring, termination,
  deletion, etc. Controlling tasks requires to deal with the
  heterogeneous batch managers and meta-schedulers that might be
  available on the infrastructure. When the infrastructure is a grid
  or a cloud, it is for instance implemented using libraries such as
  SAGA, DRMAA, OCCI and similar initiatives \todo{refs needed}.
\item Data control: operations to manage data on the infrastructure,
  such as: upload, download, deletion, browsing, replication, caching,
  etc. Data movements can be triggered by the user in the science
  gateway (\texttt{$c_1$}), to upload input data or download processed
  data. They can also be performed by the workflow engine
  (\texttt{$c_2$}), to transfer data across the infrastructure. The
  infrastructure might offer various data storage backends with
  heterogeneous interfaces. Tools and services such as
  JSAGA~\cite{reynaud2010uniform} or Data Avenue~\cite{hajnal2014data}
  can be used to homogeneize these interfaces.
\item Workflow control: operations to execute a workflow with an
  engine, including: workflow submission, monitoring, termination,
  etc. Workflow control can be coarse-grained (a.k.a. black box) or
  fine-grained (white box). In a coarse-grained model, workflow
  activities are masked and the user only has a global view of the
  workflow execution. In a fine-grained model, user is exposed to the
  workflow topology, i.e. to the outputs of the individual tasks,
  their statuses and so on. 
\item Sub-task control: operations used by tasks to submit sub-tasks
  on the infrastructure, including: submission, monitoring,
  termination, deletion, etc. Sub-task control is similar to
  interaction \texttt{b}, except that information about the parent
  task which submitted a sub-task is usually available and used for
  additional control. For instance, the parent task may wait for all
  its sub-tasks to finish before finishing, and conversely all the
  sub-tasks may be killed if the parent task is killed.
\item Pool-agent: specific to the architecture described
  in Section~\ref{sec:pool}. This is an interaction used when agents retrieve
  work from a central pool. It covers agent registration and
  de-registration to the pool, protocols to send work from the pool to
  the agent, mechanisms to update work status, and so on. 
\item Workflow conversion: translation from one workflow language to
  another one. This interaction may not be available or implementable
  for every workflow language. It has been developed mostly for
  well-structured and relatively simple workflow language such as between GWorkflowDL
  and Scufl~\cite{OLAB-09} and between the 4 languages that were involved in the
  SHIWA initiative. In SHIWA, workflow conversion is done through an
  intermediate representation, the Interoperable Workflow Intermediate
  Representation~\cite{plankensteiner-montagnat-etal:2011}, which
  allows to convert among $n$ workflow languages using $2n$
  interactions instead of $n^2$.
\end{enumerate}

@@

@@

This architecture is largely adopted, in systems such as Apache
Airvata~\cite{marru2011apache}, Vine
Toolkit~\cite{DBLP:journals/scpe/SzejnfeldDKKKKLPTWDNW10}, Virtual
Imaging Platform~\cite{GLAT-13}, the WS-PGRADE/gUSE
framework~\cite{Kacsuk2012} and the numerous science gateway instances
that rely on
WS-PGRADE/gUSE~\cite{kacsuk2014science}. Figure~\ref{fig:vip-architecture}
shows the architecture of the Virtual Imaging Platform, where the
MOTEUR workflow engine~\cite{GLAT-08i} is invoked as a service at step
\texttt{3} (interaction \texttt{d} on
Figure~\ref{fig:architectures}). Step \texttt{2} on
Figure~\ref{fig:vip-architecture} corresponds to interaction
\texttt{a$_1$}, and step \texttt{4} maps to interactions \texttt{b}
and \texttt{c$_2$} (in VIP, workflow data transfers are performed by
the tasks and embedded in their descriptions).

\begin{figure*}
\centering
\def\svgwidth{1.5\columnwidth}
\input{figures/VIP.pdf_tex}
\caption{Architecture of the Virtual Imaging Platform that integrates
  the MOTEUR workflow engine through service invocation.}
\label{fig:vip-architecture}
\end{figure*}

@@

The sub-tasking architecture is implemented in CBRAIN~\cite{SHER-14}
where it is used to integrate the PSOM workflow
engine~\cite{bellec2012pipeline} and the FSL toolkit. The CBRAIN-PSOM
integration~\cite{GLAT-16} is illustrated on
Figure~\ref{fig:cbrain-psom-architecture}. On this Figure, the science
gateway is represented by components \texttt{CBRAIN portal} and
\texttt{CBRAIN execution server} and the infrastructure consists of
\texttt{Storage servers} and \texttt{Computing cluster with shared
  file system}. Interaction \texttt{b} on
Figure~\ref{fig:architectures} is implemented by steps \texttt{4},
\texttt{5}, \texttt{7} and \texttt{8}. Interaction \texttt{c} is
implemented by steps \texttt{3} and \texttt{10}. Interaction
\texttt{e} is implemented by step \texttt{6}. The PSOM workflow engine
adopts a pilot-job architecture~\cite{turilli2015comprehensive} where
the a master coordinates workflow execution by submitting workers
and establishing direct communication channels with them. Note how
this peculiar execution model is totally supported by the sub-tasking
architecture.

@@

\begin{figure}
\def\svgwidth{\columnwidth}
\input{figures/CBRAIN-PSOM.pdf_tex}
\caption{Architecture of the CBRAIN-PSOM integration that illustrates
  the sub-tasking model (Figure extracted from~\cite{GLAT-16}). 1:
  User sends data and pipeline execution request to storage server(s)
  and CBRAIN portal. 2: CBRAIN portal sends execution request to
  execution server on cluster. 3: Execution server transfers data from
  storage server(s). 4-5: Execution server starts PSOM master via task
  scheduler. 6: PSOM master submits PSOM workers to execution
  server. 7-8: Execution server starts PSOM workers through task
  scheduler. 9: PSOM master and workers execute pipeline. 10.
  Execution server transfers results to storage server(s). \todo{Improve graphics}}
\label{fig:cbrain-psom-architecture}
\end{figure}

\subsection{Pool model}
\label{sec:pool}
See Figure~\ref{archi:agent}. Workflows are submitted by the science
gateway to a pool through interaction \texttt{d}. Agents connect to
the pool asynchronously to retrieve and execute workflows through
interaction \texttt{f}. Agents may be started according to various
policies, for instance to ensure load balancing. Agents may wrap
different types of workflow engines. Workflow engine controls tasks
and data on the infrastructure through interactions \texttt{b} and
\texttt{c$_2$}, science gateway transfers user data through
interaction \texttt{c$_1$}, and administrator installs workflow
dependencies through interaction \texttt{a$_1$}.

The pool model was implemented in the SHIWA pool~\cite{ROGE-13}, see
Figure~\ref{fig:shiwa-pool-architecture}. In Figure
~\ref{fig:shiwa-pool-architecture}, interaction \texttt{d} of
Figure~\ref{archi:agent} is implemented in 3 different types of calls
to the pool: workflow submission (\texttt{1} \& \texttt{2}), workflow
status retrieval (\texttt{a} and \texttt{b}), and workflow retrieval
(\texttt{13} \& \texttt{14}). In Figure
~\ref{fig:shiwa-pool-architecture}, interaction \texttt{f} of
Figure~\ref{archi:agent} is implemented through 2 types of calls:
workflow instance retrieval (\texttt{4}, \texttt{5} and \texttt{6}),
and workflow instance update (\texttt{11} and \texttt{12}). Workflow
instance retrieval is used by the agents to fetch work from the
pool. Workflow instance update is used by the agents to update
workflow statuses. In the SHIWA pool, agents can wrap different types
of workflow engines to execute workflows expressed with different
languages. Calls \texttt{0}, \texttt{7}, \texttt{8} and \texttt{9} are
used by workflow engine plugins to declare their supported language
and launch engines, and by workflow engines to report their status to
the agent. These calls are specific to the SHIWA Pool implementation
of the \texttt{agent} component and therefore have no corresponding
representation in Figure~\ref{archi:agent}.

@@

See Figure~\ref{archi:nested}. In nested workflows (Figure~\ref{archi:nested}-Left), an activity of a
\emph{parent} workflow executed by the \emph{parent} engine is itself
a \emph{child} workflow that is executed by a \emph{child} engine. The
parent and child engines might use different workflow description
languages. They might also execute workflows on different
infrastructures. The parent workflow is also called meta-workflow. The
science gateway communicates with the parent engine through
interaction \texttt{*$_1$}. The science gateway also communicate with
the infrastructure to transfer user data through abstract interactions
\texttt{*$_2$} and \texttt{*$_3$}. Both workflow engines communicate
with the infrastructure through abstract interactions as well,
\texttt{*$_4$} and \texttt{*$_5$}. The parent engine communicates with
the child engine through abstract interaction
\texttt{*$_6$}. Administrator installs workflows in the science
gateway through interaction \texttt{a$_2$}, and installs software on
infrastructures through interactions \texttt{a$_1$}.

Nested workflows are abstract architectural patterns that can be
instantiated in the various architectures described previously. We
focus on instantiation with the service invocation model(see
Figure~\ref{archi:nested}-Right) as this is the most used
architecture. In such an instantiation, we assume that the parent and
child workflow engines are distinct pieces of software that require
different workflow services invoked by distinct \texttt{d}
interactions. If this is not the case, then workflow services can be
collapsed in a single one with a \texttt{d} interaction with
itself. Workflow engines communicate with infrastructures using
\texttt{b} and \texttt{c$_2$}. Science gateway transfer user data to
infrastructures using \texttt{c$_1$} interactions.

Nested workflows have long been available in workflow engines, for
instance in the Taverna workbench~\cite{oinn2004taverna}. They are
also used implicitly in several platforms where workflow engines are
wrapped in workflow activities as any other command-line tool. Nested
workflows were also notably used by the SHIWA Science Gateway to
implement so-called Coarse-Grained workflow
interoperability~\cite{terstyanszky2014enabling}, i.e. to integrate
various workflow engines in a consistent
platform. Figure~\ref{fig:shiwa-architecture} shows the architecture
used in the SHIWA Science Gateway for nested workflow execution with
service invocation as represented in Figure~\ref{archi:nested}. The
parent workflow engine is WS-PGRADE, invoked as a service in the
Science Gateway (step \texttt{1} on
Figure~\ref{fig:shiwa-architecture}, interaction \texttt{d} on
Figure~\ref{archi:nested}). Ten different child engines can be used by
nested workflows, invoked through the \texttt{Submission service}
(step \texttt{2}, interaction \texttt{d}). Each of these engines can
submit jobs and transfer data to a distributed computing
infrastructure (DCI, step \texttt{3}, interactions \texttt{b} and
\texttt{c$_2$}). Data interactions (\texttt{c}) and application
porting ones (\texttt{a}) are not represented on
Figure~\ref{fig:shiwa-architecture}.

>>>>

The architectures are diagramed in Figure~\ref{fig:architectures}
using the graphical notations shown in
Figure~\ref{fig:notations}. Architectures are described from their
main software \emph{components} and \emph{interactions}. Software
components include science gateway and infrastructure in all
architectures while workflow service, workflow pool and agent are
involved in some architectures. The workflow engine itself is
represented by a specific symbol with dotted lines. Software
interactions among these components are represented with
arrows. \emph{Abstract} interactions are specific types of
interactions that may be implemented by various different software
interactions. They are represented with dotted arrows and labeled with
 \texttt{*}. An architecture that has an abstract interaction is an
abstract architecture. Red color is used to represent the components
and interactions that are specific to workflow execution, i.e. that
are not involved in any other process. Tasks refer to computing tasks
that are created by the workflow engine and executed on the
infrastructure. Data represents any type of file, or database that is
involved in the workflow execution and stored on the infrastructure
while the workflow executes. It does not cover workflow parameters
such as strings, numbers, etc. The remainder of this section describes
the different types of interactions involved in the architectures, and
details each architecture.

@@

\item Workflow integration: consists in adding a new workflow to the
  system so that users can execute it. It is triggered by an
  administrator of the science gateway and it results in an interface,
  for instance a web form, where users can enter the parameters of the
  workflow to be executed. The interaction has two aspects:
  (\texttt{a$_1$}) the programs used in the workflow are installed on
  the infrastructure, which may or may not require administrative
  privileges on the infrastructure; (\texttt{a$_2$}) the workflow is
  configured in the science gateway so that it becomes available to
  users. Note that integrating a workflow is not the same process as
  integrating a workflow \emph{engine}.
\item Task control: operations to manage tasks on the infrastructure,
  including: authentication, submission, monitoring, termination,
  deletion, etc. Controlling tasks requires to deal with the
  heterogeneous batch managers and meta-schedulers that might be
  available on the infrastructure. When the infrastructure is a grid
  or a cloud, it is for instance implemented using libraries such as
  SAGA, DRMAA, OCCI and similar initiatives \todo{refs needed}.
\item Data control: operations to manage data on the infrastructure,
  such as: upload, download, deletion, browsing, replication, caching,
  etc. Data movements can be triggered by the user in the science
  gateway (\texttt{$c_1$}), to upload input data or download processed
  data. They can also be performed by the workflow engine
  (\texttt{$c_2$}), to transfer data across the infrastructure. The
  infrastructure might offer various data storage backends with
  heterogeneous interfaces. Tools and services such as
  JSAGA~\cite{reynaud2010uniform} or Data Avenue~\cite{hajnal2014data}
  can be used to homogeneize these interfaces.
\item Workflow control: operations to execute a workflow with an
  engine, including: workflow submission, monitoring, termination,
  etc. Workflow control can be coarse-grained (a.k.a. black box) or
  fine-grained (white box). In a coarse-grained model, workflow
  activities are masked and the user only has a global view of the
  workflow execution. In a fine-grained model, user is exposed to the
  workflow topology, i.e. to the outputs of the individual tasks,
  their statuses and so on. 
\item Sub-task control: operations used by tasks to submit sub-tasks
  on the infrastructure, including: submission, monitoring,
  termination, deletion, etc. Sub-task control is similar to
  interaction \texttt{b}, except that information about the parent
  task which submitted a sub-task is usually available and used for
  additional control. For instance, the parent task may wait for all
  its sub-tasks to finish before finishing, and conversely all the
  sub-tasks may be killed if the parent task is killed.
\item Pool-agent: specific to the architecture described
  in Section~\ref{sec:pool}. This is an interaction used when agents retrieve
  work from a central pool. It covers agent registration and
  de-registration to the pool, protocols to send work from the pool to
  the agent, mechanisms to update work status, and so on. 
\item Workflow conversion: translation from one workflow language to
  another one. This interaction may not be available or implementable
  for every workflow language. It has been developed mostly for
  well-structured and relatively simple workflow language such as between GWorkflowDL
  and Scufl~\cite{OLAB-09} and between the 4 languages that were involved in the
  SHIWA initiative. In SHIWA, workflow conversion is done through an
  intermediate representation, the Interoperable Workflow Intermediate
  Representation~\cite{plankensteiner-montagnat-etal:2011}, which
  allows to convert among $n$ workflow languages using $2n$
  interactions instead of $n^2$.
\end{enumerate}

@@

\begin{figure*}
\centering
\def\svgwidth{1.5\columnwidth}
\input{figures/VIP.pdf_tex}
\caption{Architecture of the Virtual Imaging Platform that integrates
  the MOTEUR engine through service invocation.}
\label{fig:vip-architecture}
\end{figure*}

@@

This architecture is largely adopted, in systems such as Apache
Airvata~\cite{marru2011apache}, Vine
Toolkit~\cite{DBLP:journals/scpe/SzejnfeldDKKKKLPTWDNW10}, Virtual
Imaging Platform~\cite{GLAT-13}, the WS-PGRADE/gUSE
framework~\cite{Kacsuk2012} and the numerous science gateway instances
that rely on
WS-PGRADE/gUSE~\cite{kacsuk2014science}. Figure~\ref{fig:vip-architecture}
shows the architecture of the Virtual Imaging Platform, where the
MOTEUR workflow engine~\cite{GLAT-08i} is invoked as a service at step
\texttt{3} (interaction \texttt{d} on
Figure~\ref{fig:architectures}). Step \texttt{2} on
Figure~\ref{fig:vip-architecture} corresponds to interaction
\texttt{c$_1$}, and step \texttt{4} maps to interactions \texttt{b}
and \texttt{c$_2$} (in VIP, workflow data transfers are performed by
the tasks and embedded in their descriptions).

@@

The sub-tasking architecture is implemented in CBRAIN~\cite{SHER-14}
where it is used to integrate the PSOM workflow
engine~\cite{bellec2012pipeline} and the FSL toolkit. The CBRAIN-PSOM
integration~\cite{GLAT-16} is illustrated on
Figure~\ref{fig:cbrain-psom-architecture}. On this Figure, the science
gateway is represented by components \texttt{CBRAIN portal} and
\texttt{CBRAIN execution server} and the infrastructure consists of
\texttt{Storage servers} and \texttt{Computing cluster with shared
  file system}. Interaction \texttt{b} on
Figure~\ref{fig:architectures} is implemented by steps \texttt{4},
\texttt{5}, \texttt{7} and \texttt{8} (regular interactions with batch
manager). Interaction \texttt{c} is implemented by steps \texttt{3}
and \texttt{10}. Interaction \texttt{e} is implemented by step
\texttt{6}. The PSOM workflow engine adopts a pilot-job
architecture~\cite{turilli2015comprehensive} where the a master
coordinates workflow execution by submitting workers and establishing
direct communication channels with them. Note how this peculiar
execution model is totally supported by the sub-tasking architecture.

@@

\begin{figure}
\def\svgwidth{\columnwidth}
\input{figures/CBRAIN-PSOM.pdf_tex}
\caption{Architecture of the CBRAIN-PSOM integration that illustrates
  the sub-tasking model (Figure extracted from~\cite{GLAT-16}). 1:
  User sends data and workflow execution request to storage server(s)
  and CBRAIN portal. 2: CBRAIN portal sends execution request to
  execution server on cluster. 3: Execution server transfers data from
  storage server(s). 4-5: Execution server starts PSOM master via task
  scheduler. 6: PSOM master submits PSOM workers to execution
  server. 7-8: Execution server starts PSOM workers through task
  scheduler. 9: PSOM master and workers execute workflow. 10.
  Execution server transfers results to storage server(s). \todo{Improve graphics}}
\label{fig:cbrain-psom-architecture}
\end{figure}

\subsection{Pool model}
\label{sec:pool}
See Figure~\ref{archi:agent}. Workflows are submitted by the science
gateway to a pool through interaction \texttt{d}. Agents connect to
the pool asynchronously to retrieve and execute workflows through
interaction \texttt{f}. Agents may be started according to various
policies, for instance to ensure load balancing. Agents may wrap
different types of workflow engines. Workflow engine controls tasks
and data on the infrastructure through interactions \texttt{b} and
\texttt{c$_2$}, science gateway transfers user data through
interaction \texttt{c$_1$}, and administrator installs workflows
through interaction \texttt{a$_1$} and \texttt{a$_2$}.

The pool model was implemented in the SHIWA pool~\cite{ROGE-13}, see
Figure~\ref{fig:shiwa-pool-architecture}. In Figure
~\ref{fig:shiwa-pool-architecture}, interaction \texttt{d} of
Figure~\ref{archi:agent} is implemented in 3 different types of calls
to the pool: workflow submission (\texttt{1} \& \texttt{2}), workflow
status retrieval (\texttt{a} and \texttt{b}), and workflow retrieval
(\texttt{13} \& \texttt{14}). In Figure
~\ref{fig:shiwa-pool-architecture}, interaction \texttt{f} of
Figure~\ref{archi:agent} is implemented through 2 types of calls:
workflow instance retrieval (\texttt{4}, \texttt{5} and \texttt{6}),
and workflow instance update (\texttt{11} and \texttt{12}). Workflow
instance retrieval is used by the agents to fetch work from the
pool. Workflow instance update is used by the agents to update
workflow statuses. In the SHIWA pool, agents can wrap different types
of workflow engines to execute workflows expressed with different
languages. Calls \texttt{0}, \texttt{7}, \texttt{8} and \texttt{9} on
Figure~\ref{fig:shiwa-pool-architecture} are used by workflow engine
plugins to declare their supported language and launch engines, and by
workflow engines to report their status to the agent. These calls are
specific to the SHIWA Pool implementation of the \texttt{agent}
component and therefore have no corresponding representation in
Figure~\ref{archi:agent}.

@@

See Figure~\ref{archi:nested}. In nested workflows (Figure~\ref{archi:nested}-Left), an activity of a
\emph{parent} workflow executed by the \emph{parent} engine is itself
a workflow, called \emph{child} workflow, that is executed by a \emph{child} engine. The
parent and child engines might use different workflow description
languages. They might also execute workflows on different
infrastructures. The parent workflow is also called meta-workflow. The
science gateway communicates with the parent engine through
interaction \texttt{*$_1$}. The science gateway also communicate with
the infrastructure to transfer user data through abstract interactions
\texttt{*$_2$} and \texttt{*$_3$}. Both workflow engines communicate
with the infrastructure through abstract interactions as well,
\texttt{*$_4$} and \texttt{*$_5$}. The parent engine communicates with
the child engine through abstract interaction
\texttt{*$_6$}. Administrator installs workflows in the science
gateway through interaction \texttt{a$_2$}, and installs software on
infrastructures through interactions \texttt{a$_1$}.

Nested workflows are abstract architectural patterns that can be
instantiated in the various architectures described previously. We
focus on instantiation with the service invocation model (see
Figure~\ref{archi:nested}-Right) as this is the most used
architecture. In such an instantiation, we assume that the parent and
child workflow engines are distinct pieces of software that require
different workflow services invoked by distinct \texttt{d}
interactions. If this is not the case, then workflow services can be
collapsed in a single one with a \texttt{d} interaction with
itself. Workflow engines communicate with infrastructures using
\texttt{b} and \texttt{c$_2$}. Science gateway transfer user data to
infrastructures using \texttt{c$_1$} interactions.

Nested workflows have long been available in workflow engines, for
instance in the Taverna workbench~\cite{oinn2004taverna}. They are
also used implicitly in several platforms where workflow engines are
wrapped in workflow activities as any other command-line tool. Nested
workflows were notably used by the SHIWA Science Gateway to implement
so-called Coarse-Grained workflow
interoperability~\cite{terstyanszky2014enabling}, i.e. to integrate
various workflow engines in a consistent
platform. Figure~\ref{fig:shiwa-architecture} shows the architecture
used in the SHIWA Science Gateway for nested workflow execution with
service invocation as represented in Figure~\ref{archi:nested}. The
parent workflow engine is \texttt{WS-PGRADE}, invoked as a service in
the Science Gateway (step \texttt{1} on
Figure~\ref{fig:shiwa-architecture}, interaction \texttt{d} on
Figure~\ref{archi:nested}). Ten different child engines can be used by
nested workflows, invoked through the \texttt{Submission service}
(step \texttt{2}, interaction \texttt{d}). Each of these engines can
submit jobs and transfer data to a distributed computing
infrastructure (\texttt{DCI}, step \texttt{3}, interactions \texttt{b}
and \texttt{c$_2$}). Data interactions (\texttt{c}) and application
porting ones (\texttt{a}) are not represented on
Figure~\ref{fig:shiwa-architecture}.
